---
title: "Chapter 5"
author: "[Bruce Hartman](https://drbrucehartman.net/brucewebsite/)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

# Introduction
Solve Chapter 5 Problems from [the HEADS text](#r1).

# Packages

Regular R packages:
```{r}
library(tidyverse)
library(ggfortify)
library(GGally)
library(broom)
library(skimr)
library(gt)
library(patchwork)
library(yardstick)
library(car)
```

For this text:
```{r}
library(HEADS)
```

# Data
```{r}
data(Table_5_1, package="HEADS")
Table_5_1
```

# Chapter 5

## Linear Regression (5.1)

Figure 5.1.
```{r}
ggplot(Table_5_1, aes(Age, SBP)) +
  geom_point() +
  labs(y = "Systolic Blood Pressure")
```

Make predictions and add predictions to plot Figure 5.2
```{r}
fit=lm(SBP~Age, data=Table_5_1)
pred = predict(fit)
ggplot(Table_5_1, aes(Age, SBP)) +
  geom_point() +
  labs(y = "Systolic Blood Pressure") +
  geom_line(aes(y=pred), color="blue")
```

Alternative plot:
```{r}
ggplot(Table_5_1, aes(Age, SBP)) +
  geom_point() +
  labs(y = "Systolic Blood Pressure") +
  geom_smooth(method="lm", formula='y~x', se=F, color="blue")
```

```{r}
fit %>% coef
fit %>% tidy
```

### Validation of Coefficients (5.1.2)

Confidence intervals:
```{r}
fit %>% tidy(conf.int=T)
```

Hypothesis tests:

- $H0: \beta1 = 0$ vs. 
- $H1: \beta1 ̸= 0.$

```{r}
fit %>% tidy %>% dplyr::select(term, estimate, statistic, p.value)
```

R-squared:
```{r}
fit %>% glance
fit %>% anova %>% tidy()
```

## Multiple Linear Regression (5.2)

### Example 5.2.1
The dataset is in Table 5.3 and is based on [Robertson and Armitage, 1959](#r2).

```{r}
data(Table_5_3)
Table_5_3
```

Figure 5.3.
```{r fig.width=12}
p1 = ggplot(Table_5_3, aes(x=LogDose)) +
  geom_point(aes(y=RT)) +
  labs(y="Recovery Time",
       x="Log Dose",
       subtitle="Log Dose vs. Recovery Time (hours)")
p2 = ggplot(Table_5_3, aes(x=DBP)) +
  geom_point(aes(y=RT)) +
  labs(y="Recovery Time",
       x="Diastolic Blood Pressure",
       subtitle="Diastolic Blood Pressure vs. Recovery Time (hours)")

p1 + p2 + plot_annotation(tag_levels='a')
```
Simple Way:
```{r}
Table_5_3 %>% dplyr::select(LogDose,DBP,RT) %>% GGally::ggpairs(progress=F)
```
:::{#a1}
Regression and overall results:
:::
```{r}
fit53 = lm(RT~LogDose+DBP, data=Table_5_3)
fit53 %>% glance
```
:::{#a2}
Coefficients:
:::
```{r}
fit53 %>% tidy(conf.int=T)
```

:::{#a3}
We should always check residuals:
:::
```{r}
fit53 %>% autoplot(1:3)
```
We won't comment on residuals here.

### Example 5.2.2
The F-test comes from the [glance output](#a1).

### Example 5.2.3 and 5.2.4
The coefficient tests and confidence intervals come from the [tidy output](#a2).

## Variable Transformation (5.3)

### Example 5.3.1 and 5.3.2
Categorical variables can be dichotomous with just two values, and usually encoded 0-1 or with text. Some can have three or more levels, often encoded as text.

Data from Table 5.4:
```{r}
data(Table_5_4)
Table_5_4
```
Code the ethnicity using factor.
```{r}
Table_5_4$Ethnicity = as.factor(Table_5_4$Ethnicity)
```

Regression model:
```{r}
fit54 = lm(WaitingTime ~ Ethnicity, data=Table_5_4)
fit54 %>% tidy
```
Non-Latino patients get kidneys on average 303.6 days earlier.

Coding Ethnicity as -1 for Latino and 1 for Non-Latino:
```{r}
Table_5_4a = Table_5_4 %>% 
  mutate(Ethnicity1 = ifelse(Ethnicity == "Latino", 1, -1))
fit54a = lm(WaitingTime ~ Ethnicity1, data=Table_5_4a)
fit54a %>% tidy
```
Matches the book.

Another approach:
```{r}
ggplot(Table_5_4) +
  geom_boxplot(aes(x=Ethnicity, y=WaitingTime))
```


# References
:::{#r1}
Mehrotra, Sanjay, Kevin Bui, and Hari Balasubramanian (2024). Healthcare Engineering,
Analytics, and Decision Sciences: An Introduction. December 11, 2024.
(https://sites.google.com/view/healthcareengineering/)
:::

::: {#r2}
JD Robertson and P Armitage (1959) Comparison of two hypotensive agents. Anaesthesia, 14(1):53–64. 
:::

