---
title: "Chapter-2"
author: "[Bruce Hartman](https://drbrucehartman.net/brucewebsite/)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

# Introduction
Solve Chapter 2 Problems from [the HEADS text](#r1).

# Packages
Regular R packages:
```{r}
library(tidyverse)
library(ggfortify)
library(GGally)
library(broom)
library(skimr)
library(gt)
library(patchwork)
library(yardstick)
library(car)
```
Github packages:
```{r}
library(devtools)
install_github("bhartman2/mbadtools")
library(mbadtools)
```


For this text:
```{r}
library(HEADS)
```

# Data
```{r}
data(bpx_i)
bpx_i %>% skim
```

```{r}
data(trigly_i)
trigly_i %>% skim
```


# Chapter 2

# Exercise 3
Part 3b: use bpx dataset.
```{r}
tibble(
  bpx_i %>% dplyr::select(contains("SY")) %>% filter(BPXSY1>140) %>% 
    summarize(bigsy =n()),
  bpx_i %>% dplyr::select(contains("SY")) %>% filter(!is.na(BPXSY1)) %>% 
    summarize(valid=n())
) %>% mutate(ratiopct = 100*bigsy/valid)
```

Part 3c: Use trigly dataset.
```{r}
tibble(
  trigly_i %>% filter(LBDLDL>150) %>% summarize(bigldl=n()),
  trigly_i %>% filter(!is.na(LBDLDL)) %>% summarize(valid=n())
) %>% mutate(ratiopct = 100*bigldl/valid)
```
# Exercise 4
```{r}
data(HealthcareCostsMEPS)
HealthcareCostsMEPS %>% glimpse
```
Part (a)
```{r}
HealthcareCostsMEPS %>% distinct(CONDITIONCOUNT) %>% arrange(CONDITIONCOUNT)
```
CONDITIONCOUNT ranges from 0 to 12.

```{r}
df = HealthcareCostsMEPS %>% group_by(CONDITIONCOUNT) %>%
  summarize( pwt = sum(WEIGHT) ) %>% 
  mutate( prop = 100*pwt/sum(pwt))
ggplot(df) +
  geom_col(aes(x=CONDITIONCOUNT,y=prop ), fill="lightblue", color="black")+
  labs(x="Number of conditions", y="Population (percent)")
```


# References
:::{#r1}
Mehrotra, Sanjay, Kevin Bui, and Hari Balasubramanian (2024). Healthcare Engineering,
Analytics, and Decision Sciences: An Introduction. December 11, 2024.
(https://sites.google.com/view/healthcareengineering/)
:::
