---
title: "Chapter-3"
author: "[Bruce Hartman](https://drbrucehartman.net/brucewebsite/)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

# Introduction
Solve Chapter 3 Problems from [the HEADS text](#r1).

# Packages

Regular R packages:
```{r}
library(tidyverse)
library(ggfortify)
library(GGally)
library(broom)
library(skimr)
library(gt)
library(patchwork)
library(yardstick)
library(car)
```

For this text:
```{r}
library(HEADS)
```

# Data
Use the <code>birthwt</code> dataset.
```{r}
data(birthwt)
birthwt %>% skim
```

# Chapter 3

## Z-tests (3.3.1)

### Lower-tailed test
Let us test the following hypothesis for <code>bwt</code>:

- $H0 : μ = 3000;$ vs. 
- $H1 : μ < 3000$.

In order to perform the lower-tailed z-test, we compute the sample mean and sample standard deviation. The
sample mean is computed using the <code>mean()</code>
 function, and the sample standard deviation is computed using
the <code>sd()</code> function. It is done as follows, per the book:
```{r}
# textbook solution
mean_bwt <- mean(birthwt$bwt)
mean_bwt
## [1] 2944.587
#compute standard deviation
sd_bwt <- sd(birthwt$bwt)
sd_bwt
## [1] 729.2143
```

The modern way is to use <code>summarize</code> and create a tibble.
```{r}
birthwt %>% summarize(mean_bwt = mean(bwt), 
                      sd_bwt = sd(bwt), 
                      n = n()
                      )
```

$z$-statistic: add to tibble.
```{r}
birthwt %>% summarize(mean_bwt = mean(bwt), sd_bwt = sd(bwt), n = n(),
                      z = (mean_bwt - 3000)/(sd_bwt/sqrt(n)))
```
P-value:
```{r}
birthwt %>% summarize(mean_bwt = mean(bwt), 
                      sd_bwt = sd(bwt), 
                      n = n(),
                      z = (mean_bwt - 3000)/(sd_bwt/sqrt(n)),
                      p.value = pnorm(z)
                      )
```

### Upper-tailed test
New hypothesis for <code>bwt</code>:

- $H0 : μ = 2700;$ vs. 
- $H1 : μ > 2700$.

```{r}
birthwt %>% summarize(mean_bwt = mean(bwt), 
                      sd_bwt = sd(bwt), 
                      n = n(),
                      z.lower = (mean_bwt - 3000)/(sd_bwt/sqrt(n)),
                      p.lower = pnorm(z.lower),
                      z.upper = (mean_bwt-2700)/(sd_bwt/sqrt(n)),
                      p.upper = 1-pnorm(z.upper)                        
                      )
```
### Two-tailed test
New hypothesis for <code>bwt</code>:

- $H0 : μ = 2800;$ vs. 
- $H1 : μ \neq 2800$.

```{r}
birthwt %>% summarize(mean_bwt = mean(bwt), 
                      sd_bwt = sd(bwt), 
                      n = n(),
                      z.lower = (mean_bwt - 3000)/(sd_bwt/sqrt(n)),
                      p.lower = pnorm(z.lower),
                      z.upper = (mean_bwt-2700)/(sd_bwt/sqrt(n)),
                      p.upper = 1-pnorm(z.upper),
                      z.two = (mean_bwt-2800)/(sd_bwt/sqrt(n)),
                      p.two = 2*min(1-pnorm(z.two),pnorm(z.two))
                      )
```
The table shows the results of the three hypothesis tests.

### t.test Function
We can use the <code>t.test</code> function to do all three. It converges to the normal z-test when the number of observations is large enough (say, over 30 or 40). We have 189 here. 
```{r}
ht.lower = birthwt %>% dplyr::select(bwt) %>% 
  t.test(alternative="l", mu=3000) %>% tidy()

ht.upper = birthwt %>% dplyr::select(bwt) %>% 
  t.test(alternative="g", mu=2700) %>% tidy()

ht.two = birthwt %>% dplyr::select(bwt) %>% 
  t.test(alternative="t", mu=2800) %>% tidy()

ht.lower
ht.upper
ht.two
```

Comparing the z-test with the t-test:
```{r}
#add comparison to the data
result = birthwt %>% summarize(mean_bwt = mean(bwt), 
                      sd_bwt = sd(bwt), 
                      n = n(),
                      z.lower = (mean_bwt - 3000)/(sd_bwt/sqrt(n)),
                      p.lower = pnorm(z.lower),
                      z.upper = (mean_bwt-2700)/(sd_bwt/sqrt(n)),
                      p.upper = 1-pnorm(z.upper),
                      z.two = (mean_bwt-2800)/(sd_bwt/sqrt(n)),
                      p.two = 2*min(1-pnorm(z.two),pnorm(z.two))
                      ) %>% 
  mutate(t.lower = ht.lower$statistic, 
         t.upper=ht.upper$statistic,
         t.two = ht.two$statistic) %>% 
  mutate(tp.lower = ht.lower$p.value, 
         tp.upper=ht.upper$p.value,
         tp.two = ht.two$p.value)

result %>% dplyr::select(z.lower, t.lower,
         z.upper, t.upper,
         z.two, t.two)

result %>% 
  dplyr::select(p.lower, tp.lower,
         p.upper, tp.upper,
         p.two, tp.two) 
```
A perfect match for z and t statistic values. 
The lower p-values match to 3 places. 
The upper and two-sided p-values make the z-test less conservative, showing a smaller probability of the null hypothesis being true. 
The t-distribution has 'fatter tails' than the normal distribution for z.

## Two-sample z (or t) tests
For the two-sample case we use the <code>smoke</code> column, which is binary 0-1.
```{r}
birthwt %>% dplyr::select(smoke) %>% skim
```

It's better to use the built-in function <code>t.test</code>, which can also perform a two-sample t-test. Using the function for a two-sample test is similar for population means. It's easier to use the formula interface, rather than specify the two classes. We do not need to specify $\mu$ since its default value is 0- the two classes have equal means, so their difference is zero. 

We perform a two-sample t-test on the hypotheses for <code>bwt</code>:

- $H0 : μ_{non-smoker}−μ_{smoker} = 0;$ vs. 
- $H1 : μ_{non-smoker}−μ_{smoker} > 0$

```{r}
t.test(bwt ~ smoke, data=birthwt, alternative="g") %>% 
  tidy()
```


# References
:::{#r1}
Mehrotra, Sanjay, Kevin Bui, and Hari Balasubramanian (2024). Healthcare Engineering,
Analytics, and Decision Sciences: An Introduction. December 11, 2024.
(https://sites.google.com/view/healthcareengineering/)
:::
