---
title: "Chapter-5-Exercises"
author: "[Bruce Hartman](https://drbrucehartman.net/brucewebsite/)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

# Introduction
Solve Chapter 5 Exercises from [the HEADS text](#r1).

# Packages

Regular R packages:
```{r}
library(tidyverse)
library(ggfortify)
library(GGally)
library(broom)
library(skimr)
library(gt)
library(patchwork)
library(yardstick)
library(car)
```

For this text:
```{r}
library(HEADS)
```


# Exercise 15
Catheter length.
```{r}
data(Table_5_9)
Table_5_9
```

Part (a):
```{r}
fit59 = lm(CatheterLength~Height, data=Table_5_9)
fit59 %>% glance
fit59 %>% tidy()
```
Part (b):
```{r}
fit59b = lm(CatheterLength~Weight, data=Table_5_9)
fit59b %>% glance
fit59b %>% tidy()
```

Part (c):
The models have similar R-squared values. I both cases the coefficients are significant.

Part (d):
We can put both patients in the same newdata, because each regression has just one driver.
```{r}
ndata = data.frame(Height=52.7, Weight=77.5)

preda = predict(fit59, newdata=ndata, se.fit=T)
predb = predict(fit59b, newdata=ndata, se.fit=T)
tibble(
 lengtha = preda$fit,
 lengthb = predb$fit
)
```

# Exercise 16
```{r}
data(Table_5_10)
Table_5_10
```

Part (a):
```{r}
Table_5_10 %>% GGally::ggpairs(progress=F)
```
Not much relation with Weight, more with Age.

```{r}
fit510a = lm(BloodFatContent ~ Age, data=Table_5_10)
fit510aw = lm(BloodFatContent ~ Age + Weight, data=Table_5_10)
fit510a %>% glance
fit510a %>% tidy
anova(fit510a, fit510aw)
```
Age alone has an R-squared of 70%; the Age coefficient of 5.32 is significant. Adding Weight to the regression does not produce a significant improvement (p.value=.57)

# Exercise 17
```{r}
data(Table_5_11)
Table_5_11
Table_5_11 %>% GGally::ggpairs(progress=F)
fit511 = lm(DiastolicBloodPressure ~ Cholesterol, data=Table_5_11)
fit511 %>% glance
fit511 %>% tidy
```
R-squared is very low. Overall p.value of regression is .15, not significant.  Cannot justify a relation.

# Exercise 18
```{r}
data(Table_5_12)
Table_5_12
Table_5_12 %>% GGally::ggpairs(progress=F)
```
Part (a):
```{r}
fit512 = lm(SystolicBloodPressure ~ Age, data=Table_5_12)
fit512 %>% glance
fit512 %>% tidy
```
The R-squared is 63%; the pairs graph shows the relation might be significant. The tidy report shows a significant coefficient of Age is 1.41 mmHg/year.

# Exercise 19
```{r}
data(Table_5_13)
Table_5_13
```
Part (a):
We can look at this with a pairs plot, to get both a graph and correlation coefficients.
```{r}
Table_5_13 %>% GGally::ggpairs(progress=F)
```
Part (b):
```{r}
fit513 = lm(SystolicBP~BMI, data=Table_5_13)
fit513 %>% summary
```
R-squared is 52% and BMI coefficient = 1.42 is significant. There is a relation explaining about half the variation in SystolicBP.

Part (c):
We use a prediction interval because it says 'someone', not a group.
```{r}
pred = predict(fit513, newdata=data.frame(BMI=23), interval="prediction")
pred
```
The lower and upper confidence limits define the confidence interval in this prediction. 

# Exercise 20
```{r}
data(Table_5_14)
Table_5_14
```
Part (a):
```{r}
fit514 = lm(PatientSatisfaction~., data=Table_5_14)
fit514 %>% summary
```

Part (b) and (c)
```{r}
pred = predict(fit514, newdata=data.frame(LengthofStay=2, PreviousAdmissions=8), interval="p")
pred
```
The confidence interval for this patient is satisfaction between 99 and 195. Adjusted R-squared is 63%, and PreviousAdmissions is a significant coefficient, whereas LengthOfStay is not.

# Exercise 21
```{r}
data(PrescriptionsData)
```
Wrangling: set dummy for week 21 on.
```{r}
PrescriptionsData = PrescriptionsData %>% 
  mutate(MedB = case_when(
    Week >= 21 ~ 1,
    .default = 0
    ))
PrescriptionsData %>% glimpse
```
Regression:
```{r}
fit521 = lm(Prescriptions ~ ., data = PrescriptionsData)
fit521 %>% summary
fit521 %>% confint()
```

Adjusted R-squared is 71%, and both coefficients are significant. It appears introduction of MedB caused a decline of about 178 Rx of MedA with a 95% confidence interval of (-215, -142).

# Exercise 22
LogLikelihoods

# Exercise 23

```{r}
Data = data.frame(t, Prescriptions)
Data = Data %>% 
  mutate(MedB = case_when(t >= 5 ~ 1, .default = 0),
         MedG = case_when(t >= 8 ~ 1, .default = 0)
  )

formula = "Prescriptions ~ t + MedB + MedG"
fit523 = lm(formula, data=Data)
```


# Exercise 24
Table 5.16

# Exercise 25
```{r}
data("HealthcareCostsMEPS")
HealthcareCostsMEPS %>% glimpse
```
Part (a):
```{r}
fit521a = lm(HEALTHCOSTS~AGE, data=HealthcareCostsMEPS)
fit521b = lm(HEALTHCOSTS~CONDITIONCOUNT, data=HealthcareCostsMEPS)
fit521a %>% tidy(conf.int=T)
fit521b %>% tidy(conf.int=T)
```
Both AGE and CONDITIONCOUNT are significant predictors of HEALTHCOSTS, with estimates of $106 per year for AGE and $1989 per condition.

Part (b):
```{r}
fit521m = lm(HEALTHCOSTS~.-WEIGHT-DUPERSID-HEARTDISEASE, data=HealthcareCostsMEPS)
# F-test and R-squared
fit521m %>% glance
fit521m %>% tidy
fit521m %>% summary
```
The adjusted R-squared is only 13%, but at least one coefficient is significant.

Part (c): adding interactions
```{r}
fit521i = lm(HEALTHCOSTS~.-WEIGHT + SEX:RACE + FAMINC10:AGE - DUPERSID- HEARTDISEASE, data=HealthcareCostsMEPS)
fit521i %>% tidy()
fit521i %>% glance
fit521i %>% summary
```
They don't change the Adjusted R-squared very much. It actually went down. CONDITIONCOUNT and AGE are significant coefficients. Neither interaction term is significant.

Part (d): stepwise regression.
```{r}
library(leaps)
# have to omit one disease to keep from being degenerate
fit521sf = regsubsets(HEALTHCOSTS ~ . - HEARTDISEASE, 
                      data=(HealthcareCostsMEPS %>% dplyr::select(-WEIGHT, -DUPERSID)),
                      method ="forward", nvmax=24)
# summary
sum521sf = summary(fit521sf)
# examine cp, df for graph
x = data.frame(n=1:24, cp=round(sum521sf$cp, 3)); xlow=which.min(x$cp)
ggplot(x) +
  geom_line(aes(x=n, y=cp)) +
  annotate("text", x=xlow, y=x$cp[xlow], label=xlow)
# display coefficients of best fit
fit521sf %>% coef(xlow) %>% enframe(name="Driver", value="Coefficient") %>% 
  mutate(Coefficient=round(Coefficient,3))
```
```{r}
library(leaps)
# have to omit one disease to keep from being degenerate
fit521sb = regsubsets(HEALTHCOSTS ~ . - HEARTDISEASE, 
                      data=(HealthcareCostsMEPS %>% dplyr::select(-WEIGHT, -DUPERSID)),
                      method ="backward", nvmax=24)
# summary
sum521sb = summary(fit521sb)
# examine cp, df for graph
x = data.frame(n=1:24, cp=round(sum521sb$cp, 3)); xlow=which.min(x$cp)
ggplot(x) +
  geom_line(aes(x=n, y=cp)) +
  annotate("text", x=xlow, y=x$cp[xlow], label=xlow)
# display coefficients of best fit
fit521sb %>% coef(xlow) %>% enframe(name="Driver", value="Coefficient") %>% 
  mutate(Coefficient=round(Coefficient,3))
```



# Exercise 26
James Lopez Bernal, Steven Cummins, and Antonio Gasparrini. Interrupted time series regression for
the evaluation of public health interventions: a tutorial. Int J Epidemiol., 46:348–355, 2017.

# Exercise 27

# Exercise 32
```{r}
data(Covid2020)
Covid2020
```
Wrangle date:
```{r}
Covid2020 = Covid2020 %>% mutate(Date = mdy(DATE_OF_INTEREST), .after=1)
```
Plots:
```{r}
ggplot(Covid2020) +
  geom_line(aes(Date,Cases), color="blue") +
  geom_line(aes(Date,Hospitalizations), color="red") +
  geom_point(aes(Date,Deaths), color="black")
```



# References
:::{#r1}
Mehrotra, Sanjay, Kevin Bui, and Hari Balasubramanian (2024). Healthcare Engineering,
Analytics, and Decision Sciences: An Introduction. December 11, 2024.
(https://sites.google.com/view/healthcareengineering/)
:::
