---
title: "Chapter-4"
author: "[Bruce Hartman](https://drbrucehartman.net/brucewebsite/)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

# Introduction
Solve Chapter 4 Problems from [the HEADS text](#r1).

# Packages

Regular R packages:
```{r}
library(tidyverse)
library(ggfortify)
library(GGally)
library(broom)
library(skimr)
library(gt)
library(patchwork)
library(yardstick)
library(car)
```

Github packages:
```{r}
library(devtools)
install_github("bhartman2/mbadtools")
library(mbadtools)
```

For this text:
```{r}
library(HEADS)
```

# Data
```{r}
data(ControlChartConstants, package="HEADS")
```

# Chapter 4

## Control Chart Constants
This chart is prepared by simulation and may be slightly different from the one in the text.
```{r}
ControlChartConstants %>% round(digits=3)
```

# References
:::{#r1}
Mehrotra, Sanjay, Kevin Bui, and Hari Balasubramanian (2024). Healthcare Engineering,
Analytics, and Decision Sciences: An Introduction. December 11, 2024.
(https://sites.google.com/view/healthcareengineering/)
:::
