---
title: "NHANES Data"
author: "[Bruce Hartman](https://drbrucehartman.net/brucewebsite/)"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

# Introduction
Load the NHANES files for the HEADS course. 

# Packages

Regular R packages:
```{r}
library(tidyverse)
library(ggfortify)
library(GGally)
library(broom)
library(skimr)
library(gt)
library(patchwork)
library(yardstick)
library(car)
```

# Data

Download the nhanesA package as a zip file from the site in the text, since it has been banned for a violation.

This downloaded zip file package must be installed using terminal command
R CMD INSTALL "C:\Users\bruce\OneDrive\Documents\R Material\HEADS\nhanesA_1.1.tar.gz"

Following that one can make it available.
```{r}
library(nhanesA)
```

Now we get the filenames in the text (a) using the function provided.
```{r}
nhanesTables('DEMO', year=2015, details=T)
nhanesTables('EXAM', year=2015, details=T)
nhanesTables('LAB', year=2015, details=T)
```

```{r}
browseNHANES()
```


We get the files themselves using the R scripts in /data_raw.

# Data
Now we can make the files available using the <code>data()</code>
 function.
 
```{r}
data(demo_i)
data(bmx_i)
data(bpx_i)
data(trigly_i)
```

And check out the data.

The Demographic data:
```{r}
demo_i %>% skim()
```

The Examination data:
```{r}
bmx_i %>% skim()

bpx_i %>% skim()
```

The Laboratory data.
```{r}
trigly_i %>% skim()
```


# Documentation
The documentation for the NHANES data for 2015 can be accessed at [the NHANES data website](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx)

```{r}
browseURL("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/DEMO_I.htm")
browseURL("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/BMX_I.htm")
browseURL("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/BPX_I.htm")
browseURL("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2015/DataFiles/TRIGLY_I.htm")
```

# References
:::{#r1}
[NHANES data](https://wwwn.cdc.gov/nchs/nhanes/Default.aspx) Center for Disease Control, NHANES data. (https://wwwn.cdc.gov/nchs/nhanes/Default.aspx)
:::

:::{#r2}
Mehrotra, Sanjay, Kevin Bui, and Hari Balasubramanian (2024). Healthcare Engineering,
Analytics, and Decision Sciences: An Introduction. December 11, 2024.
(https://sites.google.com/view/healthcareengineering/)
:::
